CuO Analysis Details

### 2019-04-01 paper/analysis.tex
cleaned up RDM choice, need to work on rest before tomorrow

\documentclass{article}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{amsmath}
\usepackage{subcaption}
\title{CuO Analysis Details}
\author{Shivesh Pathak}
\begin{document}
\maketitle
\section{Experiment}
APES measurements in 1996 indicate that the low-lying states of the CuO molecule are composed of 3d$^9$ and 3d$^10$ states, namely 9 and 10 electrons on the copper d orbitals [ref 48.pdf].

@@

They find the ground state to be of type $X^2\Pi$ and the first excited state to be of type $Y^2\Sigma^+$, which are consistent with higher resolution PES measurements and various other theoretical and experimental measurements [ref, \textbf{clean this up}]. The $X^2\Pi \rightarrow Y^2\Sigma^+$ excitation corresponds to an excitation $O p_\sigma \rightarrow O p_\pi$, and has a measured energy of 0.97 eV. The PES measurements also find a block of excitations starting at 2.20 eV which correspond to the 3d$^9$ excited states of CuO. Earlier measurements have found similar results. Both these PES measurements and Laser Induced Flourescence measurements in 1980 [ref] have found a quartet state at 1.91 eV. This state has been assigned as $^4\Sigma$ state, but the argumentation for its assignment is quite loose and based on a simple single particle model. Shown below are the most recent APES measurements for the CuO molecule, all state assignments except for the $b$ state have been verified by other experiments.

Given that the lowest energy states are 3d10 in character and higher energy excited states at 3d9, it is informative to look at the properties of the copper atom with 10 and 9 electrons respectively. The Cu$^+$ atom can have either configurations 3d$^10$4s$^0 (^1S)$ or 3d$^9$4s$^1 (^3D, ^1D)$ [ref]. The $^3D$ is lower in energy than the $^1D$ state by about 0.54 eV, indicating that a Hund's coupling between the Cu 4s and 3d orbitals of $J_H = -0.54 eV$.

\section{Sampling scheme}
In order to do DMD we need to densely sample the low energy space. In our case here, we will define the LES as the span of some base states. Since we know that our system has a low energy $3d9$ and $3d10$ sector, we will consider base states with either 9 or 10 electrons in the d-orbitals. Further, in order to consider the effects of a possible Hund's coupling arising in our effective theory from the Cu $J_H$, we will need to consider states with both $S_z=1/2, 3/2$. When moving to QMC, we will not be able to consider linear combinations of states with different $S_z$, so in order to sample the effect of $J_H$ densely, we would also like to consider "spin-flip partner" states in the $S_z=1/2$ sector. Consider for example a state 3d$^9$2p$_{pi}^3$4s$^1$ in the $S_z=1/2$ sector. The half-filled orbitals can have electrons with configuration $up\ up\ dn$ or $up\ dn\ up$, the prior which would have a Hund's energy $\frac{1}{2}J_H$ and the latter $-\frac{1}{2}J_H$. We would also like to work with single determinant base states, since we have found that single determinant Slater-Jastrow wave functions are quite accurate for transition metal oxide molecules [ref]. In order to meet all of these considerations, we chose to collect base states by using spin and spatial-symmetry targeted unrestricted Kohn Sham eigenstates. We generated all unique base states in the 3d9 and 3d10 sectors using this method. In order to sample the span of the base states, namely our defined LES, we used a shell sampling technique which samples in shells of fixed radius around each of the base states using base state weights of 0.1 - 1.0 in increments of 0.1. Our final set of states in our span were then multiplied by a three-body Jastrow factor, which was optimized on the UKS ground state, and passed into DMC with a timestep of 0.01 and Tmoves on to calculate the 1-/2-rdms and total energy.

\section{RDM basis}
There were two choices of the RDM basis that we could have used, either an IAO basis or an MO basis. The IAO basis is more theoretically pleasing because it's a set of functions which best capture the variation between the the Cu 3d, O 2p, and Cu 4s orbitals from our different base states. However, the IAO basis has the drawback that many parameters will be correlated, for example $n_{p_\pi}$ and $t_\pi$. The MO basis choice is more arbitrary because we simply chose a set of spatially symmetric ($p_x, p_y$ look identical for example) ROKS molecular orbitals, since we don't know a way of constructing "intrinsic molecular orbitals." The MOs however yield the benefit that they are decorrelated for the most part, because to zeroth order the excitations of CuO look very similar to single-particle molecular orbital excitations. I therefore chose to do my model analysis using a mixture of both bases, namely using an MO basis for the non-interacting parts (n, t) but using an IAO basis for the interacting pieces (J, U, etc.).

>>>>

\documentclass{article}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage[margin=1.0in]{geometry}
\usepackage{lineno}
\linenumbers
\linespread{2}
\renewcommand\linenumberfont{\normalfont\bfseries\small}

\begin{document}
\section{Experiment}
Anion photoelectron spectroscopy (APES) measurements from 1996 indicate that the low-lying states of the CuO molecule are composed of 3d9 and 3d10 states [ref 48.pdf]. Shown below are the most recent APES measurements for the CuO molecule, all state assignments except for the $b^4\Sigma$ state have been verified by other experiments.

@@

The APES measurements yield a ground state of CuO is $X^2\Pi$ and the first excited state $Y^2\Sigma^+$, which are consistent with higher resolution PES measurements and various other theoretical and experimental measurements [multiple ref]. The $X^2\Pi \rightarrow Y^2\Sigma^+$ excitation corresponds to an excitation $O p_\sigma \rightarrow O p_\pi$, and has a measured energy of 0.97 eV. The APES measurements also find a block of excitations starting at 2.20 eV which correspond to the 3d9 excited states of CuO. Earlier measurements have found similar results. Further, both these APES measurements and Laser Induced Flourescence measurements in 1980 [ref] have found a quartet state at 1.91 eV. This state has been assigned as $^4\Sigma$, but the argumentation for its assignment is based on an assumption that the quartet state measured in APES is the lowest possible energy quartet state.

Further, since the lowest energy states of the CuO molecule are 3d104s0, 3d94s0 and 3d94s1 in character, it is instructive to look at experimental measurements of the properties of the Cu$^+$ and Cu$^{2+}$ ions [ref]. The Cu$^{2+}$ ion has a ground state of character 3d94s0, and a first excited state around 6.7 eV of 3d84s1, indicating that the 3d8 sector of eigenstates may be ignored in our study. The Cu$^+$ ion has a ground state character of 3d10, and two excited states at 2.72 and 3.26 eV of character 3d94s1. These two low energy excited states differ in their total spin, the lower energy state being a $^3D$ state and the higher energy state being a $^1D$ state. This indicates to us that there is a Hund's coupling on the copper atom of $J_H = -0.54 eV$.

\section{Sampling scheme}
In order to do DMD we need to densely sample the low energy space (LES). In our case, we will define the LES as the span of a set of base states. Since we know that our system has a low energy 3d9 and 3d10 sector, we will consider base states with either 9 or 10 electrons in the d-orbitals. Further, in order to consider the effects of a possible Hund's coupling arising in our effective theory from the Cu atom $J_H$, we will need to consider states with both $S_z=1/2, 3/2$. When moving to QMC, we will not be able to consider linear combinations of states with different $S_z$, so in order to sample the effect of $J_H$ densely, we would also like to consider "spin-flip partner" states in the $S_z=1/2$ sector. Consider for example a state Cu3d9\! 4s1 O2p$_{\pi}$3 in the $S_z=1/2$ sector. The half-filled orbitals can have electrons with configuration $up\ up\ dn$ or $up\ dn\ up$, the prior which would have a Hund's descriptor $\frac{1}{2}$ and the latter $-\frac{1}{2}$. Further, we would like to use single determinant Slater-Jastrow wave functions since they have proven to be quite accurate trial wave functions on previous QMC calculations for transition metal oxide molecules [ref]. In order to meet all of these considerations, we chose to collect base states by using spin and spatial-symmetry targeted unrestricted Kohn Sham (UKS) [ref] single determinant states using a B3LYP functional [ref] and Trail-Needs basis with the corresponding pseudopotentials [ref].

First, we generated all the unique base states in the 3d9 and 3d10 sectors using this method. In order to sample the span of the base states we used a shell sampling technique which samples in shells of fixed radius around each of the base states using base state weights of 0.1 - 1.0 in increments of 0.1. Our final set of states in our span were then multiplied by a three-body Jastrow factor, which was optimized on the UKS ground state. Fixed node Diffusion Monte Carlo (FN-DMC) [ref] calculations were run with these states as trial wave functions with a timestep of 0.01 and Tmoves [ref]  in order to get 1-/2-density matrix elements and total energies.

\section{RDM basis}
There were two choices of the RDM basis that we could have used, either an IAO basis or an MO basis. The IAO basis is more theoretically pleasing because it's a set of functions which best capture the variation between the the Cu 3d, O 2p, and Cu 4s orbitals from our different base states. Further, the IAO basis is beneficial for representing local interactions since they are localized orbitals. However, the IAO basis has the drawback that many 1-body parameters will be correlated, for example $n_{p_\pi}$ and $t_\pi$. The MO basis choice is more arbitrary since we don't know a way of constructing "intrinsic molecular orbitals," but the RDM elements in this basis tend to vary in a less correlated fashion. This is because, to lowest order, excitations of CuO look very similar to single-particle molecular orbital excitations, so variations in the MO descriptor space are nearly orthogonal. In order to take advantage of the benefits of both bases, I chose to do my model fitting using a mixture of both bases: an MO basis for the non-interacting parts (n, t) and an IAO basis for the interacting pieces (J, U, etc.). In the end we can rotate the MO parameters into the IAO basis, leading to a fully localized representation of our model. The IAO basis I used was constructed on the collection of active MOs (Cu 3d, O 2p and Cu 4s) from every base state. The MO basis selected was the the set of active MOs from the first excited ROKS state of CuO. This is because the single-Slater determinant ground state of CuO from ROKS breaks O$p_x, p_y$ and Cu$d_{xz}, d_{yz}$ degeneracies. These degeneracies are restored in the first excited state ROKS calculation. Later in the analysis, a weighted linear regression (WLS) is used to fit our model, and it will be shown that the WLS assigns nearly zero weight to any sampled states which have a 1-rdm trace far less than 15, the total number of electrons expected in our active space, indicating that our MO model is functioning properly even though it was seemingly arbitrarily chosen. Ultimately, one just needs to ensure that the MO basis can capture variation in the electron number occupancy without having electrons vanish between relevant excitations.
